The temporal stability in flower visitation rates is driven by the asynchrony of pollinator abundances but does not necessarily explain plant reproductive success stability.

### 2023-05-11 Manuscript/Manuscript.Rmd
acknowledgement, conflict and author contributions. Grammarly in abstract and introduction

---
title: "The temporal stability in flower visitation rates is driven by the asynchrony of pollinator abundances but does not necessarely explain plant reproductive success stability."
author: "Estefanía Tobajas^1, Virginia Domínguez-García^1, Francisco P. Molina^1, Ignasi Bartomeus^1"
date: '2022-10-26'
output:
  pdf_document:
    latex_engine: xelatex
  word_document: default
header-includes: \usepackage{float} \floatplacement{figure}{H} \usepackage{caption}
  \renewcommand{\caption}{Figure}
bibliography: library.bib
csl: journalofecol.csl
editor_options:
  markdown:
    wrap: sentence
---

\^1 Estación Biológica de Doñana (EBD-CSIC).
Sevilla, Spain.

@@

2.  Using a 5-year dataset across 13 Mediterranean shrublands and 12 plant species, we assessed the effect of pollinator richness and asynchrony as drivers of the stability of plant's visitation rate and plants reproductive success.
    In addition, we analyzed whether the effects of species richness and visitation rate on plant reproduction success were consistent among years.

3.  The stability of plant's visitation rate is driven by the asynchrony of pollinator species abundances, but the effect of species richness is only indirect, as richer communities also tend to show more asynchronous populations.
    In contrast, the stability of the plant reproductive success is not related to the stability of its pollinator visitation rates.
    A year-by-year analysis reveals that in the studied system plants are unlikely to be limited by pollinator abundances except in specific years, blurring any potential effect of visitation rates stability on the stability of reproductive success.

4.  We highlight the importance of maintaining a redundancy of pollinators that visit the same plant species, given that its presence can buffer from pollinators temporal fluctuations.
    This is particularly important for pollinators with contrasting responses to environmental fluctuations, as this creates the needed asynchrony of their populations to guarantee stable levels of flower visitation rates.
    However, we warn that more complex ecosystem functions such as reproductive success, which not only depends on pollinators but on plant competition or resource availability, might be decoupled from first order ecosystem functions (i.e., visitation rates and pollen deposition) when pollination is not a limiting factor.

@@

Global environmental changes and the consequent loss of biodiversity have highlighted the need for better understanding the role of biodiversity in ecosystem function [@Hooper2005; @Lefcheck2015; @Oliver2015; @Hong2022].
We have experimental and observational evidence that biodiversity enhances aggregated community-level properties, such as total biomass [@Cardinale2006; @Cardinale2012; @Tilman2014], but the role of biodiversity for the stability of ecosystem functioning is less explored (but see @Griffin2009; @Grman2010), specially in functions involving several trophic levels such as plant pollination [@Ebeling2008; @Winfree2009a; @Garibaldi2011; @Lazaro2022].
Globally, it is estimated that at least 78% of wild plants depend to some extent on pollination by animals [@Ollerton2011].
Understanding the mechanisms driving pollinator community stability and how this relates to the stability of ecosystem functions and services delivered (i.e. pollination) is important for predicting changes in both community structure and ecosystem function.

Temporal stability is defined as the capacity of an ecosystem to dampen environmental perturbations over time while retaining ecological functions of interest [@Ives2007; @Pimm1984].
The temporal stability of the community can be influenced by the diversity of species present and the asynchrony on the species abundance fluctuations.
A higher number of species might stabilize community level functions by increasing the range of species level potential responses to environmental changes (i.e. the portfolio effect) [@McCann2000; @Tilman2006; @Hector2010; @Xu2021], although non-significant relationships between diversity and stability have also been observed [@Hooper2005; @VanderPlas2019].
Species asynchrony represents the intrinsic differences in species response to disturbances and natural fluctuations [@Loreau2013] which has been suggested as a key driver for community temporal stability [@Loreau2008].
In this sense, higher levels of species asynchrony stabilize the overall community dynamics [@Ruijven2007; @Bluthgen2016; @Craven2018].
Species asynchrony has been found to be a factor driving positive relationships between diversity and stability [@Loreau2008; @Xu2015], and often stability is more strongly associated with the degree of asynchrony across species than with species richness itself [@Bluthgen2016; @Huang2020].
Other studies have indicated that the temporal stability of the dominant species or the asynchrony among these dominant species are important in maintaining the stability of ecosystems [@Sasaki2011; @Valencia2020].
While community temporal stability should be related to the stability of ecosystem functions and services provided by such community, there is limited evidence linking both processes on multitrophic systems.

Since the reproductive success of most plant species highly depends on pollination by animals [@Ollerton2011], the stability of pollinator communities may increase the stability of plant reproductive success.
Evidence emerging from single-years studies suggest that different factors such as species richness and abundance might be important for muti-year stability too.
For example, a greater diversity of pollinators has been shown to increase plant reproductive success as a result of complementarity in service provision arising from the variation of pollinator effectiveness across space and time of the day [@Hoehn2008; @Frund2010; @Albrecht2012; @Bluthgen2011; @Frund2013; @Magrach2021]. Pollinator species may exhibit different ranges of daily temporal activity [@Pisanty2016; @Rader2016; @Zaragoza-Trello2023] and consequently may provide pollination services at different times of the day or of the season.
In addition, higher overall visitation rates are often assumed to insure reproductive success.
However, while some studies report that higher visitation rates are correlated with higher plant reproductive outputs [@Garibaldi2013; @Kaiser-Bunbury2017, @Vazquez2005], other studies do not find such relationship [@Hegland2012; @Lundgren2013].
Interestingly, studies looking at pollen limitation show that higher visitation rates are often associated with no pollen limitation, indicating than above a certain threshold, changes in visitation rates are no longer relevant [Fig. 1; @Cosacov2008; @Williams2013].
However, the occurrence and magnitude of pollen limitation can vary among years for individual populations [@Burd1994; @Ashman2004], primarily because of variation in pollinator service delivery [@Aizen2007].
All this lines of evidence suggest that results found for single year studies should scale up to muti-year stability studies [@Garibaldi2011a; @Senapathi2021], but evidence is mostly lacking for natural plant-pollinator systems.

```{r,echo=FALSE, warning=FALSE,message=FALSE, fig.height = 8,fig.width=10,out.width="60%",fig.align='center', fig.cap =" 1. Conceptual figure showing the general relationship between visitation rate of pollinators and plant reproductive success to different species richness (R1, R2, ...). As the number of visits (V1, V2, ...) from pollinators increases, the plant reproductive success is expected to increase until than above a certain threshold, changes in visitation rates are no longer relevant to plant reproductive success."}
library(ggplot2)
library(patchwork)

####IB: ESTEFANIA, necesitamos un A y B en plots de abajo, para que la gente los reconozca. EN esos paneles yo escribiria Repreocution y Visitation /Richness entro. Y los paneles A y B necesitan explicación en la leyenda, de hecho, yo creo que ahora estan girados... o eso o estoy espeso. Lo hablamos por video.

#First dataset RED
s1 <- seq(from=0, to=100, by=0.5)
l1 = log(s1, base = 6)
d1 = data.frame(l1, s1)

#Second dataset BLUE
s2 <- seq(from=0, to=100, by=0.5)
l2 = log(s2, base = 8)
d2 = data.frame(l2, s2)

#Third dataset GREEN
s3 <- seq(from=0, to=100, by=0.5)
l3 = log(s3, base = 12)
d3 = data.frame(l3, s3)

to=ggplot(d1, aes(s1, l1)) +
  geom_line(color = "#e41a1c", linewidth=1.3) +
  geom_line(data =  d2, aes(s2,l2), color = "#377eb8",linewidth=1.3) +
  geom_line(data =  d3, aes(s3,l3), color = "#4daf4a",linewidth=1.3) +
  ylab("Reproductive success")+
  labs(x = "") +
  theme_classic()+
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank())+
  scale_x_continuous(breaks=seq(from= 0, to= 100, by = 20), labels=c("V1","V2","V3", "V4","V5","V6"))+
  theme(text = element_text(size=18,face="bold"))+
  theme(axis.ticks.x=element_blank())+
  annotate("rect", xmin = c(0, 60), xmax = c(20,80), ymin = c(0.4, 1.35), 
           ymax = c(1.75, 2.7), linetype = 1, fill="gray88", alpha=0.3, colour = "black") +
  annotate("text", x = c(98, 98,98), y = c(2.67, 2.31,1.95), label = c("R3", "R2", "R1"), size=5)+
  annotate("text", x = c(1.25, 61.3), y = c(1.62,2.58), label = c("A", "B"), size=6)+
  theme(axis.line.x = element_line(arrow = arrow()))+
  theme(axis.line.y = element_line(arrow = arrow()))

##################################
x <- seq(0, 41, by = 0.1)
y1 <- sin(x)
y1.2 <- cos(x+4.3)

da1 <- data.frame( x, y1)
da1.2 <- data.frame( x, y1.2)

R_1=ggplot(da1, aes(x, y1))+
  geom_line( size=1) +
  geom_line(data =  da1.2, aes(x, y1.2), color = "#e41a1c",size=1)+
  theme_classic() +
  xlab("Time") +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + 
  scale_y_continuous(breaks=seq(from= -1, to= 1, by = 2), labels=c("R1","R3"))+
  theme(text = element_text(size=15, face="bold"))+
  theme(axis.title.y = element_blank()) +
  annotate("text", x = 44, y = 0.4, label = "Rep", size=4,color="#e41a1c")+
  annotate("text", x = 43, y = 0, label = "R", size=4)+ 
  theme(aspect.ratio=0.5)

###################################

x2 <- seq(0, 41, by = 0.1)
y2 <- sin(x2)
y3 = sin(x2/2.5)

da2 <- data.frame(x2, y2)
da2.1 = data.frame(x2,y3)

v_1=ggplot(da2, aes(x2, y2))+
  geom_line( size=1) +
  geom_line(data =  da2.1, aes(x2,y3), color = "#e41a1c",size=1)+
  theme_classic() +
  xlab("Time")+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+ 
  scale_y_continuous(breaks=seq(from= -1, to= 1, by = 2), labels=c("V1","V2"))+
  theme(text = element_text(size=15, face="bold"))+
  theme(axis.title.y = element_blank())+
  annotate("text", x = 44, y = -0.5, label = "Rep", size=4, color="#e41a1c")+
  annotate("text", x = 42.5, y = 0, label = "V", size=4) + 
  theme(aspect.ratio=0.5)

po1=v_1/R_1

######################
x3 <- seq(0, 57, by = 0.1)
y3 <- sin(x3)
y3.1 <- cos(x3+4.3)

da3 <- data.frame(x3, y3)
da3.1 <- data.frame(x3, y3.1)

v_2=ggplot(da3, aes(x3, y3))+
  geom_line( size=1) +
  geom_line(data =  da3.1, aes(x3,y3.1), color = "#e41a1c",size=1)+
  theme_classic() +
  xlab("Time") +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + 
  scale_y_continuous(breaks=seq(from= -1, to= 1, by = 2), labels=c("V4","V5"))+
  theme(text = element_text(size=15, face="bold"))+
  theme(axis.title.y = element_blank()) +
  annotate("text", x = 61, y = 0, label = "Rep", size=4,color="#e41a1c")+
  annotate("text", x = 59, y = 0.3, label = "V", size=4)+ 
  theme(aspect.ratio=0.5)

########################################
x4 <- seq(0, 57, by = 0.1)
y4 <- sin(x4)
y5 = sin(x4/2.5)

da4 <- data.frame(x4, y4)
da5 = data.frame(x4,y5)

R_2=ggplot(da4, aes(x4, y4))+
  geom_line( size=1) +
  geom_line(data =  da5, aes(x4,y5), color = "#e41a1c",size=1)+
  theme_classic() +
  xlab("Time")+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+ 
  scale_y_continuous(breaks=seq(from= -1, to= 1, by = 2), labels=c("R1","R3"))+
  theme(text = element_text(size=15, face="bold"))+
  theme(axis.title.y = element_blank())+
  annotate("text", x = 61, y = -0.6, label = "Rep", size=4, color="#e41a1c")+
  annotate("text", x = 59, y = 0.4, label = "R", size=4)+ 
  theme(aspect.ratio=0.5)

po2=v_2/R_2

to/ (po1 |po2)
```

Here, we use a 5-year dataset collected for a total of 12 plant species across 13 Mediterranean shrubland sites.
However, to study the role of pollinator richness and asynchrony as drivers of stability of its visitation rates we only use the 5 most represented plant species across years and sites.
Further, we evaluate whether more stable pollinator communities foster a more stable plant reproductive success along years, measuring both the stability of fruit and seed set.
We expect that the stability of visitation rates will increase with higher richness and asynchrony of pollinators.
In addition, a greater stability of plant reproductive success is expected with higher stability of visitation rates.
However, given the lack of a clear relationship between pollinator and plant reproductive output stability in our dataset, we further analyzed the variability on the yearly effect of pollinator richness and visitation rates on the reproductive success of the plants.

@@

We analyze five plant species (*Cistus crispus*, *Cistus ladanifer*, *Cistus salviifolius*, *Halimium halimifolium*, and *Lavandula pedunculata*) that were recorded in at least 5 sites [(mean = 7 ± 0.71 sites)](lavandula aparece en 5 sitios, halimiun en 6, los citus estan entre 7 y 9 sitios) in a minimum of two years per site (mean = 3.25 ± 0.16 years per plant).
The stability of pollinator visitation rates was quantified as the inverse of the coefficient of variation (mean/SD) for each plant species in each plot following Tilman (1999).
This is an adequate index given that in our dataset, as in most ecological studies, the mean and the variance scale with each other.
To examine the role of richness and asynchrony of pollinator species abundances as drivers of stability of visitation rate, we calculate pollinator richness as the total number of different pollinator species that visit each plant species in each plot.
The synchrony index for each plant species in the plot was calculated following Loreau and de Mazancourt (2008), as: $$
\begingroup\Large
\phi =\frac{\sigma_{x_T}^2}{(\sum_{i=1}^S\sigma_{x_i})^2}
\endgroup
$$ where ${\sigma_{x_T}^2}$ is the variance of the community abundance of each plot and ${\sigma_{x_i}}$ is the standard deviation of species i abundance in a community with S species.
Note that we determined abundance as the total number of individuals that were observed, whereas the visitation rate is the total number of their visits.
This synchrony measurement varies from zero, when fluctuations are perfectly asynchronous, to one, when fluctuations are perfectly synchronous [@Loreau2008].
Finally, we express asynchrony as $1-\phi$.
We also calculated the stability of fruit proportion and of the seed number for each plant species in each plot following the same methods.

@@

>>>>

---
title: "The temporal stability in flower visitation rates is driven by the asynchrony of pollinator abundances but does not necessarily explain plant reproductive success stability."
author: "Estefanía Tobajas^1, Virginia Domínguez-García^2, Francisco P. Molina^2, Ignasi Bartomeus^2"
date: '2022-10-26'
output:
  word_document: default
  pdf_document:
    latex_engine: xelatex
header-includes: \usepackage{float} \floatplacement{figure}{H} \usepackage{caption}
  \renewcommand{\caption}{Figure}
bibliography: library.bib
csl: journalofecol.csl
editor_options:
  markdown:
    wrap: sentence
---

\^1 Departamento de Biología Animal, Facultad de Biología, Universidad de Salamanca, Salamanca, Spain \^2 Estación Biológica de Doñana (EBD-CSIC).
Sevilla, Spain.

@@

2.  Using a 5-year dataset across 13 Mediterranean shrublands and 12 plant species, we assessed the effect of pollinator richness and asynchrony as drivers of stability in plant visitation rates and plant reproductive success.
    In addition, we analyzed whether the effects of species richness and visitation rate on plant reproduction success were consistent across the years.

3.  The stability of plant visitation rates is driven by the asynchrony of pollinator species abundances, but the effect of species richness is only indirect, as richer communities also tend to have more asynchronous populations.
    In contrast, the stability of plant reproductive success is not related to the stability of its pollinator visitation rates.
   A year-by-year analysis reveals that plants in the system studied are unlikely to be limited by pollinator abundance except in specific years, blurring any potential effect of visitation rate stability on reproductive success stability.

4.  We highlight the importance of maintaining a redundancy of pollinators that visit the same plant species, given that their presence can buffer the temporal fluctuations of pollinators.
    This is particularly important for pollinators with contrasting responses to environmental fluctuations, as this creates the needed asynchrony of their populations to guarantee stable levels of flower visitation rates.
    However, we warn that more complex ecosystem functions such as reproductive success, which not only depends on pollinators but on plant competition or resource availability, might be decoupled from first order ecosystem functions (i.e., visitation rates and pollen deposition) when pollination is not a limiting factor.

@@

Global environmental changes and the consequent loss of biodiversity have highlighted the need for a better understanding of the role of biodiversity in ecosystem function [@Hooper2005; @Lefcheck2015; @Oliver2015; @Hong2022]. We have experimental and observational evidence that biodiversity enhances aggregated community-level properties, such as total biomass [@Cardinale2006; @Cardinale2012; @Tilman2014], but the role of biodiversity in the stability of ecosystem functioning is less explored (but see @Griffin2009; @Grman2010), specially in functions involving several trophic levels such as plant pollination [@Ebeling2008; @Winfree2009a; @Garibaldi2011; @Lazaro2022].
Globally, it is estimated that at least 78% of wild plants depend to some extent on pollination by animals [@Ollerton2011].
Understanding the mechanisms driving pollinator community stability and how this relates to the stability of ecosystem functions and services delivered (i.e., pollination) is important for predicting changes in both community structure and ecosystem function.

Temporal stability is defined as the capacity of an ecosystem to dampen environmental perturbations over time while retaining ecological functions of interest [@Ives2007; @Pimm1984]. The temporal stability of the community can be influenced by the diversity of species present and the asynchrony of species abundance fluctuations. A higher number of species might stabilize community level functions by increasing the range of species level potential responses to environmental changes (i.e. the portfolio effect) [@McCann2000; @Tilman2006; @Hector2010; @Xu2021], although non-significant relationships between diversity and stability have also been observed [@Hooper2005; @VanderPlas2019]. Species asynchrony represents the intrinsic differences in species response to disturbances and natural fluctuations [@Loreau2013] which has been suggested as a key driver for community temporal stability [@Loreau2008]. In this sense, higher levels of species asynchrony stabilize the overall community dynamics [@Ruijven2007; @Bluthgen2016; @Craven2018]. Species asynchrony has been found to be a factor driving positive relationships between diversity and stability [@Loreau2008; @Xu2015], and often stability is more strongly associated with the degree of asynchrony across species than with species richness itself [@Bluthgen2016; @Huang2020]. Other studies have indicated that the temporal stability of the dominant species, or the asynchrony among these dominant species, is important in maintaining the stability of ecosystems [@Sasaki2011; @Valencia2020].
While community temporal stability should be related to the stability of ecosystem functions and services provided by such a community, there is limited evidence linking both processes in multitrophic systems.

Since the reproductive success of most plant species highly depends on pollination by animals [@Ollerton2011], the stability of pollinator communities may increase the stability of plant reproductive success. Evidence from single-year studies suggests that different factors such as species richness and abundance may also be important for multi-year stability. For example, a greater diversity of pollinators has been shown to increase plant reproductive success as a result of complementarity in service provision arising from the variation of pollinator effectiveness across space and time of the day [@Hoehn2008; @Frund2010; @Albrecht2012; @Bluthgen2011; @Frund2013; @Magrach2021]. Pollinator species may exhibit different ranges of daily temporal activity [@Pisanty2016; @Rader2016; @Zaragoza-Trello2023] and consequently may provide pollination services at different times of the day or of the season. In addition, higher overall visitation rates are often assumed to ensure reproductive success. However, while some studies report that higher visitation rates are correlated with higher plant reproductive output [@Garibaldi2013; @Kaiser-Bunbury2017; @Vazquez2005], other studies find no such relationship [@Hegland2012; @Lundgren2013]. Interestingly, studies looking at pollen limitation show that higher visitation rates are often associated with no pollen limitation, indicating that above a certain threshold, changes in visitation rates are no longer relevant [Fig. 1; @Cosacov2008; @Williams2013]. However, the occurrence and magnitude of pollen limitation can vary from year to year for individual populations [@Burd1994; @Ashman2004], mainly because of variations in pollinator service delivery [@Aizen2007]. All these  lines of evidence suggest that results found for single year studies should scale up to muti-year stability studies [@Garibaldi2011a; @Senapathi2021], but the evidence is mostly lacking for natural plant-pollinator systems.

![ Conceptual figure showing the general relationship between the visitation rate of pollinators and plant reproductive success to different species richness (R1, R2, ...). As the number of pollinator visits increases (V1, V2, …), plant reproductive success is expected to increase until that above a certain threshold, changes in visitation rates are no longer relevant to plant reproductive success. According to this, A-square shows that temporal variation in plant reproductive success depends on variation in visitation, while B-square shows that reproductive success varies with the variation in richness.](Figures/conceptual_plot2.png){width="400"}

Here, we use a 5-year dataset collected for a total of 12 plant species across 13 Mediterranean shrubland sites. However, to study the role of pollinator richness and asynchrony as drivers of stability of its visitation rates we only use the 5 most represented plant species across years and sites. We also assess whether more stable pollinator communities foster a more stable plant reproductive success across years, measuring both fruit and seed set stability.
We expect that the stability of visitation rates will increase with higher pollinator richness and asynchrony.
In addition, greater stability of plant reproductive success is expected with higher stability of visitation rates.
However, given the lack of a clear relationship between pollinator and plant reproductive output stability in our dataset, we further analyzed the variability in the yearly effect of pollinator richness and visitation rates on plant reproductive success.

@@

We analyze five plant species (*Cistus crispus*, *Cistus ladanifer*, *Cistus salviifolius*, *Halimium halimifolium*, and *Lavandula pedunculata*) that were recorded in at least 5 sites [(mean = 7 ± 0.71 sites)](lavandula%20aparece%20en%205%20sitios,%20halimiun%20en%206,%20los%20citus%20estan%20entre%207%20y%209%20sitios) in a minimum of two years per site (mean = 3.25 ± 0.16 years per plant).
The stability of pollinator visitation rates was quantified as the inverse of the coefficient of variation (mean/SD) for each plant species in each plot following Tilman (1999).
This is an adequate index given that in our dataset, as in most ecological studies, the mean and the variance scale with each other.
To examine the role of richness and asynchrony of pollinator species abundances as drivers of stability of visitation rate, we calculate pollinator richness as the total number of different pollinator species that visit each plant species in each plot.
The synchrony index for each plant species in the plot was calculated following Loreau and de Mazancourt (2008), as: $$
\begingroup\Large
\phi =\frac{\sigma_{x_T}^2}{(\sum_{i=1}^S\sigma_{x_i})^2}
\endgroup
$$ where ${\sigma_{x_T}^2}$ is the variance of the community abundance of each plot and ${\sigma_{x_i}}$ is the standard deviation of species i abundance in a community with S species.
Note that we determined abundance as the total number of individuals that were observed, whereas the visitation rate is the total number of their visits.
This synchrony measurement varies from zero, when fluctuations are perfectly asynchronous, to one, when fluctuations are perfectly synchronous [@Loreau2008].
Finally, we express asynchrony as $1-\phi$.
We also calculated the stability of fruit proportion and of the seed number for each plant species in each plot following the same methods.

@@

# Aknowledgements

This study was supported by the XXXXX.Estefania Tobajas thanks the Bartomeus lab and Estacion Biologica de Doñana (EBD-CSIC) for hosting during the research stay.
Estefanía Tobajas was funded by a "Postdoc contract" (Ayudas Margarita Salas. Ministerio de Universidades-Universidad de Salamanca).

# Author contributions

All authors designed the idea.
Francisco P. Molina collected the data.
Estefanía Tobajas conducted the analyses with the contributions of Virginia Domínguez-García and the guidance of Ignasi Bartomeus.
Estefanía Tobajas wrote the manuscript with contributions of all authors.

# Conflict of interest

The authors declare no conflict of interest.
